A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-08 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reviewed ensemble space section

Conceptually, a statistical ensemble is a collection of outputs, taken at once, of a reproducible preparation procedure for a physical system. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (i.e. a fair coin) or over phase space (i.e. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (i.e. a positive semi-definite self-adjoint operator of trace one). An ensemble space is the collection of all possible ensembles defined in a physical theory, together with a basic mathematical structure that all physical theories must possess. The development of a full theory of ensembles goes beyond the scope of this work, and we refer to the appendix and our ongoing work (cite).

For the purpose of this paper, we will assume an ensemble space $\Ens$ is a convex subset of a Hausdorff locally convex topological vector space $V$. The topology is the physically natural topology, where each open set represents a verifiable statement. The convex structure represents statistical mixing. For example, $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture of two ensembles $\ens[a]$ and $\ens[b]$ with probability $p$ and $\bar{p} = 1 - p$. Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i}^{\infty} p_i \ens_i$ converges in the space or not. We also assume that there is a countable family of affine continuous functions $F_i : \Ens \to \mathbb{R}$ that represent statistical quantities and fully identify each ensemble. That is, $F_i(p \ens[a] + \bar{p} \ens[b]) = p F_i(\ens[a]) + \bar{p} F_i(\ens[b])$ and $\ens[a] = \ens[b]$ if and only if $F_i(\ens[a]) = F_i(\ens[b])$ for all $i$.\footnote{It is an open question whether all ensemble spaces must possess this property.} For example, we can imagine $F_i$ representing the expectation of all polynomials of position and momentum.\footnote{This requires all probability distributions to converge to zero at infinity faster than any polynomial.} Mathematically, these statistical quantities can be used to define semi-norms on the space, giving us a locally convex topology.

Additional, each ensemble space $\Ens$ is equipped with a continuous entropy function $S : \Ens \to \mathbb{R}$. The entropy of a statistical mixture satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.} The bounds come from assuming that the entropy quantifies the variability of each ensemble. Intuitively, the variability cannot decrease during mixing, and will have an upper bound when mixing two ensembles that are non-overlapping, or orthogonal. In fact, we define two ensembles $\ens[a] \ortho \ens[b]$ to be orthogonal when they mixture gives a maximal entropy increase. This does recover the notion of orthogonality in the corresponding classical and quantum vector spaces of statistical ensembles. Lastly, mixtures preserve orthogonality in the sense that mixing non-overlapping ensembles still gives non-overlapping ensembles. More precisely, $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$.

To summarize, the above mathematical structure represents the requirements that ensembles must be experimentally well-defined, allow statistical mixture, can be identified with statistical quantities and provide a well-defined entropy.\footnote{While the full physical justification of the above mathematical structure is beyond the scope of this work, it is important to us that such justification exists and this physically straight forward. This is the essence of what we call Physical Mathematics. We will summarize some of these justifications in the appendix, leaving the full and more updated discussion in our work in progress.} We now turn to the question of how we can represent both the entropy and each ensemble in terms of some measure theoretic structure.

>>>>

\textbf{Basic requirements for an ensemble space}. It is a cardinal rule in Physical Mathematics that all axioms and mathematical definitions must be grounded in physical requirements. This way we can develop mathematical frameworks that are physically well founded. As part of our larger project, we are developing a general theory of ensemble spaces, the full treatment of which is beyond the scope of this work. We limit ourselves to give the initial premises and the final mathematical characterization, without justifications and proofs.\footnote{More details are available in our ongoing work (cite)}

We start by assuming that physical laws deal with experimentally reproducible relationships, which means that they must describe relationships between statistical ensembles. Conceptually, a statistical ensemble is the collection of outputs, taken at once, of a reproducible preparation procedure for a physical system. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (e.g. a fair coin) or over phase space (e.g. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (e.g. a positive semi-definite self-adjoint operator of trace one). An ensemble space is the collection of all possible ensembles defined in a physical theory, together with a basic mathematical structure that all physical theories must possess.

An ensemble space $\Ens$ will need to satisfy a three basic requirements. First, it needs to guarantee experimental verifiability. Mathematically, the ensemble space must be a $\mathsf{T}_0$ second countable topological space, where each open set represents an experimentally verifiable statement.\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Second, it needs to allow statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$. Mathematically, this constrains the ensemble space to be a convex set.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i}^{\infty} p_i \ens_i$ converges in the space or not.} Third, an entropy function must be defined that characterizes the variability of the elements within an ensemble.

Additionally, we will require that ensembles can be identified through real valued statistical quantities. \footnote{It is an open question whether all ensemble spaces must possess this property. In a previous work we showed that the dense linear ordering require highly idealized assumptions.} A statistical quantity represents the expectation of a real valued quantity, and is therefore an affine continuous functions $F : \Ens \to \mathbb{R}$. That is, $F(p \ens[a] + \bar{p} \ens[b]) = p F(\ens[a]) + \bar{p} F(\ens[b])$. The last requirement is that there exists a family $F_i$ of statistical quantities that fully identify each ensemble. That is, $F_i(p \ens[a] + \bar{p} \ens[b]) = p F_i(\ens[a]) + \bar{p} F_i(\ens[b])$ and $\ens[a] = \ens[b]$ if and only if $F_i(\ens[a]) = F_i(\ens[b])$ for all $i$.

These requirements interact with each other leading to a rich mathematical structure that connect to many of the structures we use in mathematical physics. In this paper, we will examine only connections to measure theoretic representations.

\textbf{Ensemble spaces as convex sets.} In this work, an ensemble space $\Ens$ is a convex subset of a metrizable second countable locally convex topological vector space $V$.\footnote{The entropy constraints forces the topology to be $\mathsf{T}_1$ the convex space to embed into a vector space. A second countable $\mathsf{T}_1$ TVS is metrizable. The family of statistical quantities induce a family of semi-norms on the space, which must be countable due to second countability, which makes the TVS locally convex.} The entropy is function $S : \Ens \to \mathbb{R}$ that satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.}  Intuitively, the variability (i.e. the entropy) cannot decrease during mixing, and will have an upper bound when mixing two ensembles whose instances are always distinguishable.

\textbf{Orthogonality and separatedness.} We will say that two ensembles are orthogonal, noted $\ens[a] \ortho \ens[b]$, when their mixture gives a maximal entropy increase. This will recover the notion of orthogonality for both classical and quantum spaces. Another notion of distinction between two ensembles is in terms of the mixtures: two ensembles are separate, noted $\ens[a] \separate \ens[b]$, they cannot be expressed as a mixture of a common ensemble. That is, there is no $\ens[c]$ such that $\ens[a] = \alpha \ens[c] + \bar{\alpha} \ens[d]$ and $\ens[b] = \beta \ens[c] + \bar{\beta} \ens$ from some $\alpha, \beta \in (0,1]$ and $\ens[d],\ens \in \Ens$. In classical spaces, these two notions coincide: two ensembles are orthgonal and separate if and only if the support of the respective probability distributions overlap on a non-measure-zero set. In quantum mechanics, these two notions are different: any two pure states are separate but are not orthogonal.

To summarize, a repeatedly verifiable physical theory requires ensembles, which must be defined experimentally, allow statistical mixing and provide an entropy. Additionally, we require the existence of enough statistical quantities to identify each ensemble. Under these assumptions, an ensemble space $\Ens$ is a convex set of a metrizable second countable locally convex topological vector space $V$. We now turn to the question of how we can represent both the entropy and each ensemble in terms of some measure theoretic structure.
